--- 
title: "Overview"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  pdf_document: default
  github_document: 
    html_preview: false
---


```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(fig.path = './assets/')
```


**Why R?**

- Open source and extensible. It is also free of charge. More importantly, you (and the community) can inspect the code for any function.
-	Very popular *programing language* for statistics. "It promotes experimentation and exploration, which improves data analysis." Lots of libraries for all kinds of specialized tasks.
-	Great for visualization. Excellent packages for graphics.
-	Excellent documentation and online help.
-	A very active and helpful community.

**What is R anyway?**

If you are coming from SAS or Stata, you are better off thinking about it as a programming language and not as a statistical environment:

- Interpreted: allows direct executions of code, which makes it slower than compiled languages.
- Dynamically typed: Easy and less boiler plate, makes metaprogramming a lot easier.
- Multi-paradigm. Although most users exploit object orientation.

The extension of the R files is usually `.R`

**RStudio**

R can be downloaded from the Comprehensive R Archive Network,
[CRAN](https//cran.r-project.org). We will be using RStudio, a popular IDE. It
is important to keep in mind that R (the language) and RStudio (the GUI) are
separate things, and it is entirely possible to use different workflows with
other tools or text editors:

- `emacs` through [ESS](http://ess.r-project.org/) (what I use).
- `vim` with the [Vim-R-Plugin](http://www.vim.org/scripts/script.php?script_id=2628).
- Sublime Text.

**A few useful resources**

There is a constantly growing collection of materials available offline and
online to learn R. The
[Journal of Statistical Software](http://www.jstatsoft.org/index) and the
[Use R!](http://www.springer.com/series/6991) series from Springer regularly
publish applications of R to different domains.

A good overview for beginners is
[Learning R](http://shop.oreilly.com/product/0636920028352.do).

SAS users may find useful
[R for SAS and SPSS users](http://www.springer.com/us/book/9781461406846),
although I have never used it myself.

For the analysis of compley survey data, you may want to take a look to ["Complex
Surveys. A Guide to Analysis Using R"](http://r-survey.r-forge.r-project.org/svybook/). 

The official documentation in CRAN (The Comprehensive R Archive Network) is
simply excellent but it goes well beyond the scope of this class.

**Looking around**

RStudio offers four basic windows. 

- Console (R interpreter)
- Code, where we will write our code.
- History/Environment
- Plots/Packages/Help

**Getting help**

The documentation of R has a well-earned reputation of being excellent. It can
be accessed through the interpreter. For instance, if we wanted to get
information about what `lm` does, or what paramaters it takes or some examples
of usage, we would type:
```{r eval=FALSE}
?lm
```

The R community is very helpful and active. If you ever get stuck in a problem, the best solution is to ask in 
[StackOverflow](http://stackoverflow.com/), a very large community of
programmers using the `#r` tag. 

For obvious reasons, the language is sometimes referred to as "Rstats" (social
media and search engines, amirite?).

Within Westat, there is a growing community of users and we have a number of
resources for Q&A and sharing information or announcements. 

**About this document**

I have prepared these materials using using
[Rmarkdown](http://rmarkdown.rstudio.com/), a format that makes it easy to
create dynamic documents. The text is written in markdown, the easiest possible
markup language ("easy-to-read, easy-to-write"): `_italic_`, `**bold**`,
... but it also allows to include chunks of executable R code. It simplifies
reproducibility and it is very easy to share.

For instance, 

<p align="left">
<img src ="assets/raw-rmarkdown.png">
</p>

is rendered as:

```{r}
N <- 10
a <- 1:N
b <- a + rnorm(N)
plot(a, b); title("An ugly plot")
```

GitHub and `git` more generally deserve a class on its own (we cover it in the
Advanced R class), but here it will only be used to host the notes.
